Communicating causal effect heterogeneity

### 2024-08-20 index.qmd
Code cleaning

---
title: Communicating causal effect heterogeneity
running-head: Communicating heterogeneity
author: 
  - name: Matti Vuorre
    orcid: 0000-0001-5052-066X
    email: mjvuorre@uvt.nl
    corresponding: true
    affiliation: 
      - ref: 1
  - name: Matthew Kay
    orcid: 0000-0001-9446-0419
    affiliation:
      - ref: 2
  - name: Niall Bolger
    orcid: 0000-0001-8698-8117
    affiliation:
      - ref: 3
affiliations:
  - id: 1
    name: Tilburg University
    department: Department of Social Psychology
  - id: 2
    name: Northwestern University
    department: Computer Science and Communication Studies
  - id: 3
    name: Columbia University
    department: Department of Psychology
abstract: |
  Advances in experimental, data collection, and analysis methods have brought population variability in psychological phenomena to the fore. Yet, current practices for interpreting such heterogeneity do not appropriately treat the uncertainty inevitable in any statistical summary. Heterogeneity is best thought of as a distribution of features with a mean (average person’s effect) and variance (between-person differences). This expected heterogeneity distribution can be further summarized e.g. as a heterogeneity interval [@bolgerCausalProcessesPsychology2019]. However, because empirical studies estimate the underlying mean and variance parameters with uncertainty, the expected distribution and interval will underestimate the actual range of plausible effects in the population. Using Bayesian hierarchical models, and with the aid of empirical datasets from social and cognitive psychology, we provide a walk-through of effective heterogeneity reporting and display tools that appropriately convey measures of uncertainty. We cover interval, proportion, and ratio measures of heterogeneity and their estimation and interpretation. These tools can be a spur to theory building, allowing researchers to widen their focus from population averages to population heterogeneity in psychological phenomena.
keywords: [heterogeneity, uncertainty, variation, hierarchical model, statistics]
authornote: |
  This manuscript is not yet peer-reviewed.
bibliography: bibliography.bib
citeproc: true
csl: apa.csl
---

@@

# Output options
is_html <- knitr::is_html_output()
opts_chunk$set(
  eval = TRUE,
  cache = TRUE,
  warning = FALSE,
  error = FALSE,
  message = FALSE,
  echo = FALSE
)
options(digits = 2)
options(tinytable_tt_digits = 3)

# Plotting options
theme_set(
  theme_classic(base_size = 9) +
    theme(
      strip.text = element_text(color = "black", hjust = 0),
      strip.background = element_rect(color = NA, fill = NA),
      strip.text.x = element_text(size = rel(0.9)),
      line = element_line(linewidth = .25),
      plot.tag = element_text(size = rel(1)),
      plot.tag.location = "margin"
    )
)

@@

Our plan is as follows. First, we review established methods for estimating and communicating expected heterogeneity of causal effects in the population using an example dataset from social psychology. We then describe additional ways in which model parameters can be transformed to describe distributions of causal effects. We review the concepts and computations underlying three heterogeneity metrics: The effect's mean and standard deviation in the population; the heterogeneity interval; and the prevalence proportion. Second, we move beyond summarizing expected degrees of heterogeneity that lack information about uncertainty to describing distributions of plausible degrees of heterogeneity. Such uncertainty distributions of population feature distributions are natural components of Bayesian hierarchical models and afford efficient tools for describing distributional uncertainty. Finally, we extend these methods to compare heterogeneity across different populations using an example dataset from cognitive psychology.

@@

```{r}
#| label: data-bolgeretal
#| include: false

@@

contrasts(dat$valence) <- c(-0.5, 0.5)
contrasts(dat$valence)
```

@@

```{r}
#| label: tbl-dat1
#| tbl-cap: First six rows of example dataset 1 [@bolgerCausalProcessesPsychology2019].

@@

This equation includes two sets of parameters: The first set contains $\beta_0$ (*beta*), the intercept, and $\beta_1$, the slope or effect of valence (V). Parameters in this set do not have subscripts: In the frequentist tradition, they are considered constants---not modelled on covariates---and typically referred to as "fixed" parameters [e.g. @raudenbushHierarchicalLinearModels2002]. The second set of parameters, $\gamma_{0j}$ (*gamma*) and $\gamma_{1j}$, have the subscript $j$ to indicate that they are person-specific deviations from the average intercept and slope, respectively. That is, $\beta_0 + \gamma_{01}$ is the intercept (average reaction time) for person $j=1$. In frequentist texts, these are typically called "random" parameters because they are modeled as varying randomly according to a specified distribution. Following standard multilevel modeling assumptions, we model $\gamma_0$ and $\gamma_1$ as multivariate normal distributed:

@@

In this equation, we assume that the person-specific deviations $\gamma_0$ and $\gamma_1$ have means of zero (because the means are added to them in equation 1.2), standard deviations $\tau$ (*tau*), and a correlation $\rho$ (*rho*). Perhaps confusingly, $\tau$s and $\rho$ are also sometimes called random effects because they describe random (co)variations of the person-specific effects. To be clear, despite this naming convention they are features of the population, not of any one group or individual.

@@

```{r}
#| label: tbl-lmer
#| tbl-cap: Parameter estimates from Model 1 (ML).

@@

```{r}
#| label: descriptors
#| include: false

@@

```{r}
#| label: fig-1
#| fig-height: 2
#| fig-width: 7
#| fig-cap: Heterogeneity distribution of valence effects and various descriptions of their expected heterogeneity as estimated with Model 1. **A.** The normal density curve defined by the point estimates of the valence effect distribution's mean ($\beta_1$) and standard deviation ($\tau_1$). Shaded areas represent areas under the normal curve within 1 (dark) and 2 (light) standard deviations of the mean. **B.** The 90\% Heterogeneity Interval as represented by a line segment with arrows, and the blue shaded area. **C.** Proportion of negative valence effects in the population (blue). **D.** Proportion of valence effects in the population that are within the region of practical equivalence to zero (ROPE; blue).

# annotation definitions
valence_dist_color = "gray25"
valence_dist_samples_color = "gray65"
desc_color = "dodgerblue1"
desc_color_dark = "dodgerblue4"
zero_color = alpha("black", 0.75)

@@

fig_1_descriptors <- fig_1 +
  scale_fill_manual(values = c("grey90", alpha(desc_color, 0.45)), na.value = "white")

@@

### Interval descriptors

@@

```{r}
#| label: fig-ml-interval
#| fig-width: 4.5
#| fig-height: 2.2
#| fig-cap: "Construction of point estimates of the limits of the heterogeneity interval. **A.** While $\\textit{HI}_{90}$ can be depicted on the probability density function (PDF), **B.** its construction is easier to depict on the cumulative distribution function (CDF), $\\Phi(x; \\beta_1, \\tau_1)$. To construct a 90% heterogeneity interval, we pass 0.05 and 0.95 to the inverse of the CDF: $\\textit{HI}_{90} = \\Phi^{-1}\\left([0.05, 0.95]; \\beta_1, \\tau_1 \\right)$."

@@

# construction of HI_90 on the CDF
fig_ml_cdf_interval <- fig_ml_cdf +
  geom_segment(
    aes(x = I(0), xend = quantile(dist_ml, 0.05), y = 0.05, yend = 0.05),
    color = desc_color
  ) +
  geom_segment(
    aes(x = quantile(dist_ml, 0.05), xend = quantile(dist_ml, 0.05), y = 0.05, yend = 0),
    color = desc_color
  ) +
  geom_segment(
    aes(x = I(0), xend = quantile(dist_ml, 0.95), y = 0.95, yend = 0.95),
    color = desc_color
  ) +
  geom_segment(
    aes(x = quantile(dist_ml, 0.95), xend = quantile(dist_ml, 0.95), y = 0.95, yend = 0),
    color = desc_color
  ) +
  geom_segment(
    aes(x = quantile(dist_ml, 0.05), xend = quantile(dist_ml, 0.95), y = 0.025, yend = 0.025),
    color = desc_color,
    arrow = arrowhead(ends = "both")
  ) +
  geom_text(
    aes(x = mu1n, y = .04),
    label = expression(bold(HI)[bold("90")]),
    data = data.frame()[1,],
    size = 3,
    parse = TRUE,
    color = desc_color,
    vjust = 0
  ) +
  annotate("text",
    x = xmin,
    y = c(0.05, 0.95),
    vjust = c(0.3, 0.7),
    label = c(0.05, 0.95),
    color = desc_color,
    hjust = 1.1,
    fontface = "bold",
    size = 2.5
  ) +
  scale_x_valence() +
  coord_cartesian(clip = "off", ylim = c(0,1))

@@

To calculate a heterogeneity interval, we first specify the desired probability limits: 
for a $\pi\%$ interval, we use the limits $(1 \pm \pi)/2$. Thus, for a 90% interval, we use 0.05 and 0.95, which together define the central 90% of the distribution. Then, we pass those limits and the estimated mean and standard deviation to the normal quantile function $\Phi^{-1}$ (*phi*, `qnorm()` in R), to get the interval: $\textit{HI}_{90} = \Phi^{-1}([.05, .95]; \beta_1, \tau_1) = \Phi^{-1}([.05, .95];$ `r mu1[1]`, `r sd1[1]`) = [`r hi90`]. In words, this function calculates the 0.05 and 0.95 quantiles of the normal distribution defined by the point estimates of the mean ($\beta_1$) and standard deviation ($\tau_1$): We expect 90% of valence effects in the population to fall in the [`r hi90`] interval. We illustrate this interval in @fig-ml-interval.

### Proportion descriptors

The above HI summarizes where a given proportion of individuals' effects in the population are. In contrast, some applications might find it more informative to summarize proportions of effects above or below some critical value, or within some critical range. For example, we might ask "What proportion of individuals in the population respond faster to positively valenced words?" In other words, we ask a question of *prevalence*: What proportion of the heterogeneity distribution is below zero? We label this quantity $p^-$ for proportion of population with negative effects.

@@

zero_line = geom_vline(
  xintercept = 0,
  linetype = "22",
  color = zero_color,
  alpha = 0.25
)

# construction of p^- on the CDF
fig_ml_cdf_pneg <- fig_ml_cdf +
  zero_line +
  geom_segment(
    aes(x = 0, xend = 0, y = 0, yend = cdf(dist_ml, 0)),
    color = desc_color
  ) +
  geom_segment(
    aes(x = 0, xend = I(0), y = cdf(dist_ml, 0)),
    color = desc_color,
    arrow = arrowhead()
  ) + 
  annotate("text", 
    x = 0, y = I(0),
    label = "0", 
    color = desc_color, 
    vjust = 1.35, 
    fontface = "bold", 
    size = 3
  ) +
  geom_text( 
    aes(x = I(0), y = cdf(dist_ml, 0)),
    label = expression(bold(p ^ "−")), 
    color = desc_color, 
    hjust = 1.2,
    vjust = 0.6,
    fontface = "bold", 
    size = 3
  ) +
  scale_x_valence(breaks = c(-0.4, -0.2, 0.2)) +
  coord_cartesian(clip = "off", ylim = c(0,1))

@@

To answer, we pass zero (the critical value) and the estimated mean and standard deviation to the normal cumulative distribution function ($\Phi$; `pnorm()` in R): 
$p^{-} = Pr(\textrm{Valence effect} \le 0) = \Phi(0; \beta_1, \tau_1) = \Phi(0;$ `r mu1[1]`, `r sd1[1]`) = `r p_less_zero`. This number is the probability that a random slope from this population would take a negative value, or, in other words, the proportion of individuals in the population with negative valence effects. We illustrate this probability in @fig-1 C and its construction using the CDF in @fig-ml-pneg.

@@

rope_line = geom_vline(
  xintercept = c(-0.1, 0.1),
  linetype = "22",
  color = zero_color,
  alpha = 0.25
)

# construction of p^- on the CDF
fig_ml_cdf_prope <- fig_ml_cdf +
  rope_line +
  geom_segment(
    aes(x = -0.1, xend = -0.1, y = 0, yend = cdf(dist_ml, -0.1)),
    color = desc_color
  ) +
  geom_segment(
    aes(x = -0.1, xend = I(0), y = cdf(dist_ml, -0.1)),
    color = desc_color,
    arrow = arrowhead()
  ) + 
  geom_segment(
    aes(x = 0.1, xend = 0.1, y = 0, yend = cdf(dist_ml, 0.1)),
    color = desc_color
  ) +
  geom_segment(
    aes(x = 0.1, xend = I(0), y = cdf(dist_ml, 0.1)),
    color = desc_color,
    arrow = arrowhead()
  ) + 
  geom_segment(
    aes(x = I(0.09), xend = I(0.09), y = cdf(dist_ml, -0.1), yend = cdf(dist_ml, 0.1)),
    color = desc_color,
    arrow = arrowhead(ends = "both")
  ) +
  annotate("text", 
    x = -0.1, y = I(0),
    label = "−ε",
    color = desc_color, 
    vjust = 1.35, 
    fontface = "bold", 
    size = 3
  ) +
  annotate("text", 
    x = 0.1, y = I(0),
    label = "+ε", 
    color = desc_color, 
    vjust = 1.35, 
    fontface = "bold", 
    size = 3
  ) +
  geom_text( 
    aes(x = I(0.12), y = mean(cdf(dist_ml, c(-0.1, 0.1))[[1]])),
    label = expression(bold(p ^ "ROPE")), 
    color = desc_color, 
    hjust = 0,
    vjust = 0.6,
    fontface = "bold", 
    size = 3
  ) +
  scale_x_valence() +
  coord_cartesian(clip = "off", ylim = c(0,1))

(
  (fig_1_prope + rope_line) +
  fig_ml_cdf_prope
) &
  theme(axis.title.y = element_text(angle = 90)) &
  plot_annotation(tag_levels = "A")
```

For example, let us imagine that a theory states that valence effects in the interval [-0.1, 0.1] are practically equivalent to zero. To calculate, we can again use the normal cumulative distribution function to calculate the proportion of individuals in the population whose valence effect falls within this interval or region of practical equivalence: $p^{ROPE} = Pr(-0.1 \le \textrm{Valence effect} \le 0.1) = \Phi(0.1; \beta_1, \tau_1) - \Phi(-0.1; \beta_1, \tau_1)$ = `r p_rope`. In words, `r p_rope` of the population is expected to have valence effects that are practically equivalent to zero. Note that this statement's validity critically depends on the chosen interval's theoretical validity. We visualize this probability in @fig-ml-prope.

### Ratio descriptors

@@

The expected normal distribution of valence effects and its transformations (@fig-1) ignore uncertainty inherent in the estimated parameters. That is, we calculated $\textit{HI}_{90}$, $p^-$, and the other heterogeneity measures from the point estimates $\beta_1$ = `r mu1[1]` and $\tau_1$ = `r sd1[1]`. We did not use any information about the precision, or uncertainty, with which these parameters were estimated. We have now arrived at the crux of the current work: How should we estimate and describe heterogeneity in psychological phenomena such that the fundamental uncertainty in the estimated parameters is retained?

@@

```{r echo = is_html, include = is_html}
#| label: ppcheck-1
#| fig-cap: "Graphical check of Model 1 performance and posterior predictive density."

@@

```{r}
#| label: tbl-samples-1
#| tbl-cap: Random draws from the posterior distributions of $\beta_1$, $\tau_1$, and their ratio.

@@

In practice, one obtains (for example) 4,000 samples from the posterior distribution using Markov Chain Monte Carlo algorithms [e.g. @standevelopmentteamStanModelingLanguage2023] through accessible software [e.g., @burknerBrmsPackageBayesian2017], and then summarizes them using familiar data processing techniques [e.g. @dplyr2023; @rcoreteamLanguageEnvironmentStatistical2023]. Here, we used the R package brms [@burknerBrmsPackageBayesian2017; @burknerAdvancedBayesianMultilevel2018] to specify the model and then draw random samples from its posterior distribution. The MCMC estimation algorithm completed in about 5 seconds on a modern laptop. We then assessed the estimation algorithm convergence graphically and numerically, and model adequacy using a graphical posterior predictive check [@gelmanBayesianDataAnalysis2013]. (We present these and other details in our supplementary online analyses.)

@@

```{r}
#| label: tbl-fit-1
#| tbl-cap: Parameter estimates from Model 1 (Bayes).

@@

Armed with the joint Bayesian posterior distribution of all model parameters (@tbl-fit-1), we can now return to the question of the distribution of valence effects in the population. We now have 4,000 draws from the posterior distribution of the heterogeneity distribution. Within each draw from the posterior distribution, we can perform any calculation we did previously on only the point estimate of the heterogeneity distribution.

We first recompute the expected heterogeneity distribution from @fig-1 using the posterior mean values of $\beta_1$ and $\tau_1$ in @fig-2 A (thick black curve). Superimposed on that normal density curve are heterogeneity distributions calculated from 100 random posterior draws of $\beta_1$ and $\tau_1$. From these curves we can see that the true distribution of valence effects might well be less or more heterogeneous than is suggested by the expectation (point estimates). We can perform the same exercise on the CDF as well (@fig-2 B).

@@

p_pdf_samples <- p_samples +
  aes(y = pdf) +
  scale_x_valence() +
  scale_y_continuous(
    breaks = extended_breaks(7),
    expand = expansion(c(0.005, 0.005)),
  ) +
  geom_rug(
    data = as_tibble(coef(fit_1)$person[,,"valence1"]) |>
      rownames_to_column("person"),
    aes(x = Estimate, y = 0), length = unit(4, "pt"), linewidth = 0.25
  ) +
  ylab("Probability density")

@@

Above, we described the heterogeneity interval as a range of values where a specific percentage of the population's slopes are expected to fall (e.g. $\textit{HI}_{90}$ for a 90% heterogeneity interval). However, a single interval cannot accommodate the uncertainty with which the underlying parameters are estimated. To carry uncertainty forward from model parameters to the $\textit{HI}_{90}$, we redo the calculations from above, but instead of using only the mean's and standard deviation's point estimates, we repeat the calculations for each of the 4,000 randomly sampled pairs of $\beta_1$ and $\tau_1$. Consequently, we get 4,000 draws from the posterior distribution of $\textit{HI}_{90}$ (@fig-cdfs-intervals).

>>>>

---
title: Communicating causal effect heterogeneity
running-head: Communicating heterogeneity
author: 
  - name: Matti Vuorre
    orcid: 0000-0001-5052-066X
    email: mjvuorre@uvt.nl
    corresponding: true
    affiliation: 
      - ref: 1
  - name: Matthew Kay
    orcid: 0000-0001-9446-0419
    affiliation:
      - ref: 2
  - name: Niall Bolger
    orcid: 0000-0001-8698-8117
    affiliation:
      - ref: 3
affiliations:
  - id: 1
    name: Tilburg University
    department: Department of Social Psychology
  - id: 2
    name: Northwestern University
    department: Computer Science and Communication Studies
  - id: 3
    name: Columbia University
    department: Department of Psychology
abstract: |
  Advances in experimental, data collection, and analysis methods have brought population variability in psychological phenomena to the fore. Yet, current practices for interpreting such heterogeneity do not appropriately treat the uncertainty inevitable in any statistical summary. Heterogeneity is best thought of as a distribution of features with a mean (average person’s effect) and variance (between-person differences). This expected heterogeneity distribution can be further summarized e.g. as a heterogeneity interval [@bolgerCausalProcessesPsychology2019]. However, because empirical studies estimate the underlying mean and variance parameters with uncertainty, the expected distribution and interval will underestimate the actual range of plausible effects in the population. Using Bayesian hierarchical models, and with the aid of empirical datasets from social and cognitive psychology, we provide a walk-through of effective heterogeneity reporting and display tools that appropriately convey measures of uncertainty. We cover interval, proportion, and ratio measures of heterogeneity and their estimation and interpretation. These tools can be a spur to theory building, allowing researchers to widen their focus from population averages to population heterogeneity in psychological phenomena.
keywords: [heterogeneity, uncertainty, variation, multilevel model, statistics, visualization]
authornote: |
  This manuscript is not yet peer-reviewed.
bibliography: bibliography.bib
citeproc: true
csl: apa.csl
---

@@

# Output options
is_html <- knitr::is_html_output()
opts_chunk$set(
  eval = TRUE,
  cache = TRUE,
  warning = FALSE,
  error = FALSE,
  message = FALSE,
  echo = is_html
)
options(digits = 2)
options(tinytable_tt_digits = 3)

# Plotting options
plot_font <- ifelse(is_html, "Fira Sans", "Libertinus Sans")
update_geom_defaults("text", list(family = plot_font))
theme_set(
  theme_classic(base_size = 10, base_family = plot_font) +
    theme(
      strip.text = element_text(color = "black", hjust = 0),
      strip.background = element_rect(color = NA, fill = NA),
      strip.text.x = element_text(size = rel(0.9)),
      line = element_line(linewidth = .25),
      plot.tag = element_text(size = rel(1)),
      plot.tag.location = "margin"
    )
)

@@

Our plan is as follows. First, we review established methods for estimating and communicating expected heterogeneity of causal effects in the population using an example dataset from social psychology. We then describe additional ways in which model parameters can be transformed to describe distributions of causal effects. We review the concepts and computations underlying three heterogeneity metrics: The effect's mean and standard deviation in the population; the heterogeneity interval; and the prevalence proportion. Second, we move beyond summarizing expected degrees of heterogeneity that lack information about uncertainty to describing distributions of plausible degrees of heterogeneity. Such uncertainty distributions of population feature distributions are natural components of Bayesian hierarchical models and afford efficient tools for describing distributional uncertainty. Finally, we extend these methods to compare heterogeneity across different populations using an example dataset from cognitive psychology. We present the computational notebook supporting this manuscript as an online supplement at {{<meta site-url>}}.

@@

```{r}
#| label: data-bolgeretal

@@

contrasts(dat$valence) <- c(-0.5, 0.5)
```

@@

```{r}
#| label: tbl-dat1
#| tbl-cap: "First six rows of example dataset 1 [@bolgerCausalProcessesPsychology2019]."

@@

This equation includes two sets of parameters: The first set contains $\beta_0$ (*beta*), the intercept, and $\beta_1$, the slope or effect of valence (V). Parameters in this set do not have subscripts: In the frequentist tradition, they are considered constants---not modelled on covariates---and typically referred to as "fixed" parameters [e.g. @raudenbushHierarchicalLinearModels2002]. The second set of parameters, $\gamma_{0j}$ (*gamma*) and $\gamma_{1j}$, have the subscript $j$ to indicate that they are person-specific deviations from the average intercept and slope, respectively. That is, $\beta_0 + \gamma_{01}$ is the intercept (average log reaction time) for person $j=1$. In frequentist texts, these are typically called "random" parameters because they are modeled as varying randomly according to a specified distribution. Following standard multilevel modeling assumptions, we model $\gamma_0$ and $\gamma_1$ as multivariate normal distributed:

@@

In this equation, we assume that the person-specific deviations $\gamma_0$ and $\gamma_1$ have means of zero (because the means are added to them in #eq-m1-2), standard deviations $\tau$ (*tau*), and a correlation $\rho$ (*rho*). Perhaps confusingly, $\tau$s and $\rho$ are also sometimes called random effects because they describe random (co)variations of the person-specific effects. To be clear, despite this naming convention they are features of the population, not of any one individual.

@@

```{r}
#| label: tbl-lmer
#| tbl-cap: "Parameter estimates from Model 1 (ML)."

@@

```{r}
#| label: descriptors

@@

```{r}
#| label: fig-1
#| fig-height: 2
#| fig-width: 7
#| fig-cap: "Heterogeneity distribution of valence effects and various descriptions of their expected heterogeneity as estimated with Model 1. **A.** The normal density curve defined by the point estimates of the valence effect distribution's mean ($\\beta_1$) and standard deviation ($\\tau_1$). Shaded areas represent areas under the normal curve within 1 (dark) and 2 (light) standard deviations of the mean. **B.** The 90\\% Heterogeneity Interval as represented by a line segment with arrows, and the blue shaded area. **C.** Proportion of negative valence effects in the population (blue). **D.** Proportion of valence effects in the population that are within the region of practical equivalence to zero (ROPE; blue)."

# annotation definitions
valence_dist_color <- "gray25"
valence_dist_samples_color <- "gray65"
desc_color <- "dodgerblue1"
desc_color_dark <- "dodgerblue4"
zero_color <- alpha("black", 0.75)

@@

fig_1_descriptors <- fig_1 +
  scale_fill_manual(
    values = c("grey90", alpha(desc_color, 0.45)), 
    na.value = "white"
  )

@@

## Interval descriptors

@@

```{r}
#| label: fig-ml-interval
#| fig-width: 4.5
#| fig-height: 2.2
#| fig-cap: "Construction of point estimates of the limits of the heterogeneity interval. While $\\textit{HI}_{90}$ can be depicted on the probability density function (PDF; **A.**), its construction is easier to depict on the cumulative distribution function (CDF; **B.**), $\\Phi(x; \\beta_1, \\tau_1)$. To construct a 90% heterogeneity interval, we pass 0.05 and 0.95 to the inverse of the CDF: $\\textit{HI}_{90} = \\Phi^{-1}\\left([0.05, 0.95]; \\beta_1, \\tau_1 \\right)$."

@@

# construction of HI_90 on the CDF
fig_ml_cdf_interval <- fig_ml_cdf +
  geom_segment(
    aes(x = I(0), xend = quantile(dist_ml, 0.05), y = 0.05, yend = 0.05),
    color = desc_color
  ) +
  geom_segment(
    aes(
      x = quantile(dist_ml, 0.05), xend = quantile(dist_ml, 0.05), 
      y = 0.05, yend = 0
    ),
    color = desc_color
  ) +
  geom_segment(
    aes(x = I(0), xend = quantile(dist_ml, 0.95), y = 0.95, yend = 0.95),
    color = desc_color
  ) +
  geom_segment(
    aes(
      x = quantile(dist_ml, 0.95), xend = quantile(dist_ml, 0.95), 
      y = 0.95, yend = 0
    ),
    color = desc_color
  ) +
  geom_segment(
    aes(
      x = quantile(dist_ml, 0.05), xend = quantile(dist_ml, 0.95), 
      y = 0.025, yend = 0.025
    ),
    color = desc_color,
    arrow = arrowhead(ends = "both")
  ) +
  geom_text(
    aes(x = mu1n, y = .04),
    label = expression(bold(HI)[bold("90")]),
    data = data.frame()[1, ],
    size = 3,
    parse = TRUE,
    color = desc_color,
    vjust = 0
  ) +
  annotate("text",
    x = xmin,
    y = c(0.05, 0.95),
    vjust = c(0.3, 0.7),
    label = c(0.05, 0.95),
    color = desc_color,
    hjust = 1.1,
    fontface = "bold",
    size = 2.5
  ) +
  scale_x_valence() +
  coord_cartesian(clip = "off", ylim = c(0, 1))

@@

To calculate a heterogeneity interval, we first specify the desired probability limits: 
for a $\pi\%$ interval, we use the limits $(1 \pm \pi)/2$. Thus, for a 90% interval, we use 0.05 and 0.95, which together define the central 90% of the distribution. Then, we pass those limits and the estimated mean and standard deviation to the normal quantile function $\Phi^{-1}$ (*phi*, `qnorm()` in R), to get the interval: $\textit{HI}_{90} = \Phi^{-1}((1 \pm \pi)/2; \beta_1, \tau_1) = \Phi^{-1}([.05, .95];$ `r mu1[1]`, `r sd1[1]`) = [`r hi90`]. In words, this function calculates the 0.05 and 0.95 quantiles of the normal distribution defined by the point estimates of the mean ($\beta_1$) and standard deviation ($\tau_1$): We expect 90% of valence effects in the population to fall in the [`r hi90`] interval. We illustrate this interval in @fig-ml-interval.

## Proportion descriptors

The above HI summarizes where a given proportion of individuals' effects in the population are. In contrast, some applications might find it more informative to summarize proportions of effects above or below some critical value, or within some critical range. For example, we might ask "What proportion of individuals in the population endorse faster to positively valenced words?" In other words, we ask a question of *prevalence*: What proportion of the heterogeneity distribution is below zero? We label this quantity $p^-$ for proportion of population with negative effects.

@@

zero_line <- geom_vline(
  xintercept = 0,
  linetype = "22",
  color = zero_color,
  alpha = 0.25
)

# construction of p^- on the CDF
fig_ml_cdf_pneg <- fig_ml_cdf +
  zero_line +
  geom_segment(
    aes(x = 0, xend = 0, y = 0, yend = cdf(dist_ml, 0)),
    color = desc_color
  ) +
  geom_segment(
    aes(x = 0, xend = I(0), y = cdf(dist_ml, 0)),
    color = desc_color,
    arrow = arrowhead()
  ) + 
  annotate(
    "text", 
    x = 0, y = I(0),
    label = "0", 
    color = desc_color, 
    vjust = 1.35, 
    fontface = "bold", 
    size = 3
  ) +
  geom_text( 
    aes(x = I(0), y = cdf(dist_ml, 0)),
    label = expression(bold(p ^ "−")), 
    color = desc_color, 
    hjust = 1.2,
    vjust = 0.6,
    fontface = "bold", 
    size = 3
  ) +
  scale_x_valence(breaks = c(-0.4, -0.2, 0.2)) +
  coord_cartesian(clip = "off", ylim = c(0, 1))

@@

To answer, we pass zero (the critical value) and the estimated mean and standard deviation to the normal cumulative distribution function ($\Phi$; `pnorm()` in R): 
$p^{-} = Pr(\textrm{Valence effect} \le 0) = \Phi(0; \beta_1, \tau_1) = \Phi(0;$ `r mu1[1]`, `r sd1[1]`) = `r p_less_zero`. This number is the probability that a random slope from this population would take a negative value, or, in other words, the proportion of individuals in the population who are expected to endorse positive words faster than negative words. We illustrate this probability and its construction using the CDF in @fig-ml-pneg.

@@

rope_line <- geom_vline(
  xintercept = c(-0.1, 0.1),
  linetype = "22",
  color = zero_color,
  alpha = 0.25
)

# construction of p^- on the CDF
fig_ml_cdf_prope <- fig_ml_cdf +
  rope_line +
  geom_segment(
    aes(x = -0.1, xend = -0.1, y = 0, yend = cdf(dist_ml, -0.1)),
    color = desc_color
  ) +
  geom_segment(
    aes(x = -0.1, xend = I(0), y = cdf(dist_ml, -0.1)),
    color = desc_color,
    arrow = arrowhead()
  ) + 
  geom_segment(
    aes(x = 0.1, xend = 0.1, y = 0, yend = cdf(dist_ml, 0.1)),
    color = desc_color
  ) +
  geom_segment(
    aes(x = 0.1, xend = I(0), y = cdf(dist_ml, 0.1)),
    color = desc_color,
    arrow = arrowhead()
  ) + 
  geom_segment(
    aes(
      x = I(0.09), xend = I(0.09), 
      y = cdf(dist_ml, -0.1), yend = cdf(dist_ml, 0.1)
    ),
    color = desc_color,
    arrow = arrowhead(ends = "both")
  ) +
  annotate("text", 
    x = -0.1, y = I(0),
    label = "−ε",
    color = desc_color, 
    vjust = 1.35, 
    fontface = "bold", 
    size = 3
  ) +
  annotate("text", 
    x = 0.1, y = I(0),
    label = "+ε", 
    color = desc_color, 
    vjust = 1.35, 
    fontface = "bold", 
    size = 3
  ) +
  geom_text( 
    aes(x = I(0.12), y = mean(cdf(dist_ml, c(-0.1, 0.1))[[1]])),
    label = expression(bold(p ^ "ROPE")), 
    color = desc_color, 
    hjust = 0,
    vjust = 0.6,
    fontface = "bold", 
    size = 3
  ) +
  scale_x_valence() +
  coord_cartesian(clip = "off", ylim = c(0, 1))

((fig_1_prope + rope_line) + fig_ml_cdf_prope) &
  theme(axis.title.y = element_text(angle = 90)) &
  plot_annotation(tag_levels = "A")
```

For example, let us imagine that a theory states that valence effects in the interval [-0.1, 0.1] are practically equivalent to zero. To calculate, we can again use the normal cumulative distribution function to calculate the proportion of individuals in the population whose valence effect falls within this interval or region of practical equivalence: $p^{ROPE} = Pr(-0.1 \le \textrm{Valence effect} \le 0.1) = \Phi(0.1; \beta_1, \tau_1) - \Phi(-0.1; \beta_1, \tau_1)$ = `r p_rope`. In words, `r p_rope` of the population is expected to have valence effects that are practically equivalent to zero. Note that this statement's validity critically depends on the chosen interval's theoretical validity. We visualize this probability, and how it is constructed using the CDF, in @fig-ml-prope.

## Ratio descriptors

@@

The expected normal distribution of valence effects and its transformations ignore uncertainty inherent in the estimated parameters. That is, we calculated $\textit{HI}_{90}$, $p^-$, and the other heterogeneity measures from the point estimates $\beta_1$ = `r mu1[1]` and $\tau_1$ = `r sd1[1]`. We did not use any information about the precision, or uncertainty, with which these parameters were estimated. We have now arrived at the crux of the current work: How should we estimate and describe heterogeneity in psychological phenomena such that the fundamental uncertainty in the estimated parameters is retained?

@@

```{r include = is_html}
#| label: ppcheck-1
#| fig-height: 2.4
#| fig-cap: "Graphical check of Model 1 performance and posterior predictive density."

@@

```{r}
#| label: tbl-samples-1
#| tbl-cap: "Random draws from the posterior distributions of $\\beta_1$, $\\tau_1$, and their ratio."

@@

In practice, one obtains (for example) 4,000 samples from the posterior distribution using Markov Chain Monte Carlo algorithms [e.g. @standevelopmentteamStanModelingLanguage2023] through accessible software [e.g., @burknerBrmsPackageBayesian2017], and then summarizes them using familiar data processing techniques [e.g. @dplyr2023; @rcoreteamLanguageEnvironmentStatistical2023]. Here, we used the R package brms [@burknerBrmsPackageBayesian2017; @burknerAdvancedBayesianMultilevel2018] to specify the model and then sample random draws from its posterior distribution. The MCMC estimation algorithm completed in about 5 seconds on a modern laptop. We then assessed the estimation algorithm convergence graphically and numerically, and model adequacy using a graphical posterior predictive check [@gelmanBayesianDataAnalysis2013]. (We present these and other details in our [online supplement](https://mvuorre.github.io/heterogeneity-uncertainty/).)

@@

```{r}
#| label: tbl-fit-1
#| tbl-cap: "Parameter estimates from Model 1 (Bayes)."

@@

Armed with the joint Bayesian posterior distribution of all model parameters, we can now returned to the question of the distribution of valence effects in the population. We have 4,000 draws from the posterior distribution of the heterogeneity distribution. Within each draw from the posterior distribution, we can perform any calculation we did previously on only the point estimate of the heterogeneity distribution.

We first recompute the expected heterogeneity distribution from @fig-1 using the posterior mean values of $\beta_1$ and $\tau_1$ in @fig-2 A (thick black curve). Superimposed on that normal density curve are heterogeneity distributions calculated from 100 random posterior draws of $\beta_1$ and $\tau_1$. From these curves we can see that the true distribution of valence effects might well be less or more heterogeneous than is suggested by the expectation (point estimates). We then perform the same exercise on the CDF as well (@fig-2 B).

@@

p_pdf_samples <- p_samples +
  aes(y = pdf) +
  scale_x_valence() +
  scale_y_continuous(
    breaks = extended_breaks(7),
    expand = expansion(c(0.005, 0.005)),
  ) +
  geom_rug(
    data = as_tibble(coef(fit_1)$person[, , "valence1"]) |>
      rownames_to_column("person"),
    aes(x = Estimate, y = 0), length = unit(4, "pt"), linewidth = 0.25
  ) +
  ylab("Probability density")

@@

Above, we described the heterogeneity interval as a range of values where a specific percentage of the population's slopes are expected to fall (e.g. $\textit{HI}_{90}$ for a 90% heterogeneity interval). However, a single interval cannot accommodate the uncertainty with which the underlying parameters are estimated. To carry uncertainty forward from model parameters to the $\textit{HI}_{90}$, we repeat the calculations from above, but instead of using only the mean's and standard deviation's point estimates, we redo the calculations for each of the 4,000 randomly sampled pairs of $\beta_1$ and $\tau_1$. Consequently, we get 4,000 draws from the posterior distribution of $\textit{HI}_{90}$ (@fig-cdfs-intervals).
